Latent Variable Clustering for Classification of Events in the AT-TPC

### 2020-04-17 doc/Manuscript/article.tex
context and motivation on pre-trained  + paragraph on clustering on latent

\subsection{Data}
 In this section we give a brief overview of the data, for a more in-depth consideration we refer to \cite{Mittig2015}, \cite{Suzuki2012} and  \cite{Bradt2017a}.

@@

\subsection{Data processing}

@@

\subsubsection{Pre-trained neural networks}

To affect the clustering of AT-TPC events, we first considered the results from \citet{Kuchera2019}. In their work, the authors build a very successful classification approach using a pre-trained neural network to preprocess the events. From their results, it is clear that the pre-trained network captures salient information about the difference in event types. Our approach builds on previous work by considering the output from the pre-trained model as a representation on which we perform clustering with a simple K-means algorithm.

In the Machine Learning community, it is not uncommon to publish packaged models with fitted parameters from image recognition contests. These models are trained on datasets with millions of images and classify between hundreds of distinct classes; one such is the imagenet dataset. In their work, \citet{Kuchera2019} used the VGG16 architecture trained on imagenet to classify AT-TPC. Following on their work we will be applying the same VGG16 model to the clustering task.

The VGG16 network is one of six analogous networks proposed by \citet{Simonyan2014}. They were runners up in the ISLVRC(ImageNet large scale visual recognition competition) of 2014 \cite{Russakovsky2015}, narrowly being beat by GooLeNet. The network architectures are fairly straightforward; for VGG16, there are sixteen layers in the network. The first thirteen of which are convolutional layers with exclusively $3 \times 3$ kernels. The choice of the kernel size is based on the fact that a stacked $3 \times 3$ kernel is equivalent to larger kernels in terms of the receptive field of the output. Three $3 \times 3$ kernels with stride $1$ have a $7 \times 7$ receptive field, but the larger kernel has $81\%$ more parameters and only one non-linearity \cite{Simonyan2014}. Stacking the smaller kernels then contributes to a lower computational cost. Additionally, there is a regularizing effect from the lowered number of parameters and increased explanatory power from the additional non-linearities. The full architecture is detailed in appendix \ref{tab:vgg}.

\subsubsection{Deep clustering: Mixture of autoencoders}\label{sec:mixae}

@@

In addition to the results presented in this section, we performed clustering with a number of different algorithms included in the \lstinline{scikit-learn} package. None of them provided any notable differences from the K-means results or were significantly worse. Notably, the DBSCAN algorithm failed to provide any useful clustering results. We find this important as one of the significant drawbacks of K-means, and the deep clustering algorithms presented in section \ref{sec:deep_clustering}, is that they are all dependent on pre-determining the number of clusters. This is not the case for DBSCAN. 
\subsection{Autoencoder clustering}

@@

\appendix  
\section{Neural network architectures}

\begin{table}
\caption[vgg architectures]{Showing the details of the VGG network architectures. Network D trained on the ImageNet \cite{Russakovsky2015} dataset the network known as  VGG16 and is what we use in this thesis.}\label{tab:vgg}
\includegraphics[width=\textwidth]{plots/vgg_architectures.png}
\end{table}

>>>>

\subsection{Data}
\todo{equalize to square plots - shorten down (maybe look at 3d)}
In this section we give a brief overview of the data, for a more in-depth consideration we refer to \cite{Mittig2015}, \cite{Suzuki2012} and  \cite{Bradt2017a}.

@@

\section{Data processing}

@@

\subsection{Pre-trained neural networks}

To affect the clustering of AT-TPC events, we first considered the results from \citet{Kuchera2019}. In their work, the authors consider the segmentation of reactions in the $^{46}$Ar experiment as a classification problem. Given ground-truth labels, they apply convolutional neural networks (CNN) to the problem. Using a CNN was in large part inspired by results from high-energy physics and, in particular, the work by \citet{Aurisano2016}.

Using a CNN algorithm comes with the caveat that the data should be image-like. This caveat is especially stringent when applying pre-trained models to new data. For the $^{46}$Ar data we project the 3D point clouds in each event to a 2D matrix representation of the X-Y plane, summing over the Z-axis. Applying a pre-trained model is a common technique in computer science, and \citet{Kuchera2019} showed that a pre-trained CNN model outperforms bespoke models for the classification of $^{46}$Ar data. We build on the classification result in this work by using a pre-trained CNN model as a spring-board for our clustering models.

In the Machine Learning community, it is not uncommon to publish packaged models with fitted parameters from image recognition contests. These models are trained on datasets with millions of images and classify between hundreds of distinct classes; one such is the imagenet dataset. In their work, \citet{Kuchera2019} used the VGG16 architecture trained on imagenet to classify ${}^{46}$Ar data. In our work we also found the VGG16 architecture to capture the strongest event-separating representation for clustering purposes.

The VGG16 network is one of six analogous networks proposed by \citet{Simonyan2014}. The network architectures are fairly straightforward; for VGG16, there are sixteen layers in the network. The first thirteen of which are convolutional layers with exclusively $3 \times 3$ kernels.  The full architecture is detailed in appendix \ref{app:vgg}.

\subsubsection{Clustering on latent spaces}

In contrast with the classification work by \citet{Kuchera2019}, we do not append a classifying layer to the VGG16 network given that we do not have access to the ground truth labels at training time in the clustering regime. We instead consider the output from the convolutional blocks of the VGG16 network as a representation of the events on which we will apply a clustering algorithm.

The clustering algorithm we found to give the strongest result is the K-means algorithm. This algorithm is contingent on having a distance measurement between two datapoints. We found that the standard choice of a euclidian norm on the vector differences provided the best outcomes.

\subsection{Deep clustering: Mixture of autoencoders}\label{sec:mixae}

@@

In addition to the results presented in this section, we performed clustering with a number of different algorithms included in the \lstinline{scikit-learn} package. None of them provided any notable differences from the K-means results or were significantly worse. Notably, the DBSCAN algorithm failed to provide any useful clustering results. We find this important as one of the significant drawbacks of K-means, and the deep clustering algorithms presented in section \ref{sec:deep_clustering}, is that they are all dependent on pre-determining the number of clusters. This is not the case for DBSCAN. 
\subsection{Autoencoder clustering}
\todo{look at grayscale images or colorbar with propto charge}

@@

\appendix  
\section{VGG16}\label{app:vgg}
The choice of the kernel size is based on the fact that a stacked $3 \times 3$ kernel is equivalent to larger kernels in terms of the receptive field of the output. Three $3 \times 3$ kernels with stride $1$ have a $7 \times 7$ receptive field, but the larger kernel has $81\%$ more parameters and only one non-linearity \cite{Simonyan2014}. Stacking the smaller kernels then contributes to a lower computational cost. Additionally, there is a regularizing effect from the lowered number of parameters and increased explanatory power from the additional non-linearities.
\begin{table}
\caption[vgg architectures]{Showing the details of the VGG network architectures. Network D trained on the ImageNet \cite{Russakovsky2015} dataset the network known as  VGG16 and is what we use in this thesis.}\label{tab:vgg}
\includegraphics[width=\textwidth]{plots/vgg_architectures.png}
\end{table}
